### 2021-04-12 paper.Rmd
use American spelling

Data commonly arrives with more than two measured variables, which makes it more complicated to plot on a page. With multiple variables, especially if there is some association between variables, this would be called multivariate or high-dimensional data. When the variables are all numeric, or quantitative, visualisation often relies on some form of dimension reduction. This can be done by taking linear projections, for example, principal component analysis [@Hotelling1933-of] or linear discriminant analysis [@Fisher1936]. It is also common to reduce dimension nonlinearly with techniques like multidimensional scaling (MDS) [@Kruskal1964-do] or t-Distributed Stochastic Neighbour Embedding (t-SNE) [@Van_der_Maaten2008-qa].

@@

The focus of this review will be on visualising high-dimensional numerical data using linear projections, in particular, as provided by the grand tour [@Asimov1985-xr; @Buja2005-cx]. The reason being that it is not feasible to adequately review the very large area of visualising high-dimensions, and there have been numerous developments in tours recently. An overview of the technique, and new modifications is provided, along with how these techniques can be used in a variety of applications, and to help understand nonlinear dimension reduction results.

While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualising structure in high dimensional data. An early technique proposed for assessing pairwise relationships between variables is the scatter plot matrix (SPLOM) [@Tukey1981-ov; @Chambers1983-dh; @Tukey1983-fj; @Carr1984-xw; @Becker1987-dw]. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables. Similarly, parallel coordinates plots (PCP) can be used to explore correlation and multicollinearity [@Inselberg1985-kf; @Wegman1990-dy]. By placing multiple variables side by side in a PCP higher order structure like clustering or lower dimensional embeddings is revealed, however the ordering of variables along the axis changes what can be learned. Another display that relies on variable ordering is the heatmap, which is widely used to visualise cluster structure in bioinformatics. @Wilkinson2009-pj provides a comprehensive history of this display in the social and natural sciences.

>>>>

Data commonly arrives with more than two measured variables, which makes it more complicated to plot on a page. With multiple variables, especially if there is some association between variables, this would be called multivariate or high-dimensional data. When the variables are all numeric, or quantitative, visualisation often relies on some form of dimension reduction. This can be done by taking linear projections, for example, principal component analysis [@Hotelling1933-of] or linear discriminant analysis [@Fisher1936]. It is also common to reduce dimension with nonlinear techniques like multidimensional scaling (MDS) [@Kruskal1964-do] or t-Distributed Stochastic Neighbor Embedding (t-SNE) [@Van_der_Maaten2008-qa].

@@

The focus of this review will be on visualizing high-dimensional numerical data using linear projections, in particular, as provided by the grand tour [@Asimov1985-xr; @Buja2005-cx]. The reason being that it is not feasible to adequately review the very large area of visualizing high-dimensions, and there have been numerous developments in tours recently. An overview of the technique, and new modifications is provided, along with how these techniques can be used in a variety of applications, and to help understand nonlinear dimension reduction results.

While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualizing structure in high dimensional data. An early technique proposed for assessing pairwise relationships between variables is the scatter plot matrix (SPLOM) [@Tukey1981-ov; @Chambers1983-dh; @Tukey1983-fj; @Carr1984-xw; @Becker1987-dw]. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables. Similarly, parallel coordinates plots (PCP) can be used to explore correlation and collinearity [@Inselberg1985-kf; @Wegman1990-dy]. By placing multiple variables side by side in a PCP higher order structure like clustering or lower dimensional embeddings are revealed, however the ordering of variables along the axis changes what can be learned. Another display that relies on variable ordering is the heatmap, which is widely used to visualise cluster structure in bioinformatics. @Wilkinson2009-pj provides a comprehensive history of this display in the social and natural sciences.
